\bf Calendar-based graphics for visualizing people's daily schedules

### 2019-10-06 reviews/response.Rmd

> Page 11 line 1. Write "For example in Figure 6..."

Done.

> Page 12. Say why polar transformation is useful.

We added "It is most useful in exhibiting cyclical patterns in the data." at the end of paragraph.

@@

> Page 17, figure 10. I don't see negative correlation. Just two slopes, both positive. Am I missing something?

TODO

> Page 19, first sentence. The meaning of this is not clear. Rephrase.

Done. We rephrased to "The previous calendar plots are static, made with **ggplot2**."

@@

> p7 "Each month conforms to a layout of 5 rows and 7 columns, where the top left is Monday of week 1, and the bottom right is Sunday of week 5 by default."  This is not true for six-week months.

TODO

> p7 "Each variable is scaled to have values in [0,1], using the minimum and maximum of all the data values to be displayed, assuming fixed scales."  Can the scale be set by the user?  Starting at 0, not minimum, can be useful.

Users can scale the original data before passing to the `frame_calendar()`.

@@

> p21 Figure 12 is an interesting possibility, but not well implemented.  The repetition of the scales in unhelpful.  0 12 24 looks like 240 12 240.  There is too much labelling and too little information.

TODO

> The case study is interesting.  Why are the data for household 2 different (fewer possible values)?

Data is of the same time lengths (and the same amount of data points) across four households. Different households have different lifestyles.

@@

> p22 "with noticeable variations on Thursdays" based on only 26 data points is overinterpretation.

It's half-hourly data, meaning 26 Thursdays * 48 per day = 1248 data points.

### 2019-10-10 reviews/response.Rmd

We thank the reviewers for their helpful feedback which has improved the readability of the paper. The point by point description of changes are below: the reviewers' comments are in red and our response is in black.

# Associate editor {-}

> What if the data's collected over different time zones?

In R, a vector (a column in a data frame) only allows for one time zone. Users need to adjust these time zones to a unified one so that data over different periods can be comparable.

@@

> Line 36, what are M and N? You explain them later, but should be at first usage.

Done.

> Page 8. I agree with reviewer 1. These figures are not needed. Move to supplementary material.

We argue to go against this comment. These figures serve as the graphical explanations of Equations (1) and (2). Readers can easily and quickly grasp the layout idea by skimming through these figures, without diving into the body.

@@

Done.

# Reviewer 1 {-}

> "Would it be possible to have a plot for each day in a calendar plot with thick lines separating the months?  That kind of plot is common for calendar heatmaps."  I think that is important.  Perhaps they can build it in.  If not, why not?

The lines separating the months are thicker than the ones separating the days in the calendar plot. The white space between months also helps to emphasise the separations between months.

@@

> p10, l50: "which occurs to May and October." -> "which occurs in May and October."

Done.

> p11, l5: "The date argument specifies the date variable used to construct the calendar layout." Not very informative. Either omit or provide an example, e.g., date in which format (numerical, text, ...), from-to dates, etc.

We elaborated a bit on this. The `date` variable tells the range of dates plotted in the calendar.

### 2019-10-10 calendar-vis.Rmd
addressed a bit of reviwer 1
more responses added

The algorithm of transforming data for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where the top left is Monday of week 1, and the bottom right is Sunday of week 5 by default. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

@@

It is assumed that the `data` is in tidy format [@wickham2014tidy], and `x`, `y` are the variables that will be mapped to the horizontal and vertical axes in each cell. For example in Figure \@ref(fig:fs-2016), the `x` is the time of the day, and `y` is the count. The `date` argument specifies the date variable used to construct the calendar layout.

>>>>

The algorithm of transforming data for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where the top left is Sunday of week 1, and the bottom right is Monday of week 5 by default. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

@@

It is assumed that the `data` is in tidy format [@wickham2014tidy], and `x`, `y` are the variables that will be mapped to the horizontal and vertical axes in each cell. For example in Figure \@ref(fig:fs-2016), the `x` is the time of the day, and `y` is the count. The `date` argument specifies the date variable in the data, facilitating the range of dates plotted in the calendar layout.

### 2019-10-10 calendar-vis.tex
more responses added

\bigskip
\begin{abstract}
Calendars are broadly used in society to display temporal information and events. This paper describes a new calendar display for plotting data, that includes a layout algorithm with many options, and faceting functionality. The functions use modular algebra on the date variable to restructure the data into a calendar format. The user can apply the grammar of graphics to create plots inside each calendar cell, and thus the displays synchronize neatly with \textbf{ggplot2} graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, is insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to events such as work days, weekends, holidays, and special events. The functions for the calendar algorithm are available in the R package \textbf{sugrrants}.
\end{abstract}

@@

A new method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout is developed. The format is created using modular arithmetic, giving a restructuring of the data, that can then be integrated into a data pipeline. The core component of the pipeline is to visualize the resulting data using the grammar of graphics \citep{wilkinson2006grammar, wickham2009ggplot2}, as used in \textbf{ggplot2} \citep{R-ggplot2}, where plots are defined as a functional mapping from variables in the data to graphical elements. The data restructuring approach is consistent with the tidy data principles available in the \textbf{tidyverse} suite of tools \citep{R-tidyverse}. The methods are implemented in a new R package called \textbf{sugrrants} \citep{R-sugrrants}.

The purpose of the calendar-based visualization is to provide insights into people's daily schedules, relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia \citep{ped}. There are many sensors installed across the inner-city area, that count pedestrians every hour (Figure \ref{fig:ped-map}). Data from 43 sensors in 2016 is analyzed here. This data can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. Patterns relative to special events (such as public holidays and recurring cultural/sporting events) would be worth studying in comparison to regular days, but the conventional displays of time series data may bury this detail.

@@

\caption{Google map of the Melbourne city area, grey dots indicate all the sensor locations. The three locations highlighted will be analyzed in the paper: the State Library is a public library; Flagstaff Station is a train station, closed on non-work days; Birrarung Marr is an outdoor park hosting many cultural and sports events.}\label{fig:ped-map}
\end{figure}

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There are some existing works in this area. For example, \citet{VanWijkCluster1999} developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts week days and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. \citet{Jones-heatmap}, \citet{R-TimeProjection}, \citet{R-ggTimeSeries}, and \citet{R-ggcal} implemented some variants of calendar-based heatmaps in R packages: \textbf{TimeProjection}, \textbf{ggTimeSeries}, and \textbf{ggcal} respectively. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Color-encoding is also low on the hierarchy of optimal variable mapping \citep{cleveland1984graphical, lam2007overview}.

The proposed algorithm goes beyond the calendar-based heatmap. The approach is developed with three conditions in mind: (1) to display time-of-day variation in addition to longer temporal components such as day-of-week and day-of-year; (2) to incorporate line graphs and other types of glyphs into the graphical toolkit for the calendar layout; (3) to enable patterns related to special events more easily pop-up to viewers. The proposed algorithm has been implemented in the \texttt{frame\_calendar()} and \texttt{facet\_calendar()} functions in the \textbf{sugrrants} package using R.

@@

The algorithm of transforming data for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data \citep{Wickham2012glyph}. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where the top left is Monday of week 1, and the bottom right is Sunday of week 5 by default. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

\begin{itemize}
\tightlist
\item
  \(k = 1, \dots , 7\) is the day of the week, that is the first day of the month.
\item
  \(d = 28, 29, 30\) or \(31\) representing the number of days in any month.
\item
  \((i, j)\) is the grid position where \(1 \le i \le 5\), is week within the month, \(1 \le j \le 7\), is day of the week.
\item
  \(g = k, \dots,(k+d)\) indexes the day in the month, inside the 35 possible cells.
\end{itemize}

The grid position for any day in the month is given by
\begin{equation}
  \begin{aligned}
  i &= \lceil (g \text{ mod } 35) / 7\rceil, \\
  j &= g \text{ mod } 7.
  \end{aligned}
  \label{eq:grid}
\end{equation}

\noindent Figure \ref{fig:month-diagram} illustrates this \((i, j)\) layout for a month where \(k=5\).

@@

To create the layout for a full year, \((m, n)\) denotes the position of the month arranged in the plot, where \(1 \le m \le M\) and \(1 \le n \le N\); \(b\) denotes the small amount of white space between each month for visual separation. Figure \ref{fig:year-diagram} illustrates this layout where \(M = 3\) and \(N = 4\).

@@

\caption{The calendar plot of hourly foot traffic at Flagstaff Station, using line glyphs. The disparities between week day and weekend along with public holiday, are immediately apparent. The arrangement of the data into a \(3 \times 4\) monthly grid represents all the traffic in 2016. Note that, the algorithm wraps the last few days in the sixth week to the top row of each month block for a compact layout, which occurs to May and October.}\label{fig:fs-2016}
\end{figure}

Figure \ref{fig:fs-2016} shows the line glyphs framed in the monthly calendar over the year 2016. This is achieved by the \texttt{frame\_calendar()} function, which computes the coordinates on the calendar for the input data variables. These can then be plotted using the usual \textbf{ggplot2} R package \citep{R-ggplot2}. Thus, the grammar of graphics can be applied.

In order to make calendar-based graphics more accessible and informative, reference lines dividing each cell and block, as well as labels indicating week day and month are also computed before plot construction.

@@

The month labels located on the top left using \((\min{(x)}, \max{(y)})\) for every \((m, n)\). The week day texts are uniformly positioned on the bottom of the whole canvas, that is \(\min{(y)}\), with the central position of a cell \(x / 2\) for each \(j\).

@@

It is assumed that the \texttt{data} is in tidy format \citep{wickham2014tidy}, and \texttt{x}, \texttt{y} are the variables that will be mapped to the horizontal and vertical axes in each cell. For example, the \texttt{x} is the time of the day, and \texttt{y} is the count (Figure \ref{fig:fs-2016}). The \texttt{date} argument specifies the date variable used to construct the calendar layout.

@@

The start of the week for a monthly calendar is adjustable. The default is Monday (\texttt{1}), which is chosen from the data perspective. The week, however, can begin with Sunday (\texttt{7}) as commonly used in the US and Canada, or other week day, subject to different countries and cultures.

@@

When \texttt{polar\ =\ TRUE}, a polar transformation is carried out on the data. The computation is similar to the one described in \citet{Wickham2012glyph}. This produces star glyphs \citep{chambers2017graphical}, where time series lines are transformed in polar coordinates, embedded in the monthly calendar layout.

@@

\caption{Line glyphs on the calendar format showing hourly foot traffic at Flagstaff Station, scaled individually by day. The shape on a single day becomes more distinctive, as compared to Figure \ref{fig:fs-2016}.}\label{fig:fs-free}
\end{figure}

The \texttt{free\_wday} scales each week day together. It can be useful to compare trends across week days, allowing relative patterns for weekends versus week days to be examined. Similarly, the \texttt{free\_mday} uses free scaling for any day within a given month.

@@

\caption{Line charts, embedded in the \(6 \times 2\) monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasized.}\label{fig:facet}
\end{figure}

@@

Almost any type of plot can be shown in a calendar pane. Most of the full range of plotting capabilities in \textbf{ggplot2} is available. An example is shown in Figure \ref{fig:scatterplot}: the panes contain lag scatterplots, constructed with local scaling for each day at Flagstaff Station, where the lagged hourly count is assigned to the \texttt{x} argument and the current hourly count to the \texttt{y} argument. It indicates strong autocorrelation on weekends, and weak autocorrelation on work days. The V-shape in the week day graphs arises when the next hour sees a substantial increase or decrease in counts. This is due to the morning and afternoon commuter patterns in Figure \ref{fig:fs-2016}: the adjacent hours are positively correlated when approaching to the peak hour but negatively correlated when moving away from the peak.

@@

\caption{Side-by-side boxplots of hourly counts for all the 43 sensors in December 2016, with the loess smooth line superimposed on each day. It shows the hourly distribution in the city as a whole. The increased variability is notable on the last day of December as New Year's Eve approaches. The month and week day are labeled in Chinese, which demonstrates the support for languages other than English.}\label{fig:chn}
\end{figure}

@@

As a data restructuring tool, the interactivity of calendar-based displays can be easily enabled, as long as the interactive graphics system remains true to the spirit of the grammar of graphics, for example, \textbf{plotly} \citep{plotly} in R. As a standalone display, an interactive tooltip can be added to show labels when mousing over a point in the calendar plot, for example the hourly count with the time of day. It is difficult to sense the values from the static display, but the tooltip makes it possible. Options in the \texttt{frame\_calendar()} function can be ported to a form of selection button or text input in a graphical user interface like R shiny \citep{R-shiny}. The display will update on the fly accordingly, via clicking or text input, as desired.

@@

Figure \ref{fig:dow} shows the energy use across days of week using boxplots. Inspecting the medians across households tells us that household 3, a family size of one couple and two children, uses more energy over the week days than other households. The relatively larger boxes for household 2 indicate greater variability in daily energy consumption with noticeable variations on Thursdays, and much higher usage over the weekends. The other two households (1 and 4) tend to consume more energy with more variation on the weekends relative to the week days, reflecting work and leisure patterns.

@@

\caption{Boxplots of daily energy usage against day of week for four households. Suggested by the medians, household 3 uses more energy than the others on the week days. By contrast, household 2 sees considerably larger variability.}\label{fig:dow}
\end{figure}

Figure \ref{fig:hod} shows energy consumption against time of day, separately by week day and weekend. Household 1 is an early riser, starting their day before 6am and going back home around 6pm on week days. They switch air conditioning on when they get home from work and keep it operating until midnight, evident from the small horizontal cluster of points around 0.8 kWh. On the other hand, the stripes above 1 kWh for household 2 indicates that air conditioning may run continuously for some periods, consuming twice the energy as household 1. A third peak occurs around 3pm for household 3 only, likely coinciding when the children arrive home from school. They also have a consistent energy pattern between week days and weekends. As for household 4, their home routine starts after 6pm on week days. Figures \ref{fig:dow} and \ref{fig:hod}, part of a traditional graphical toolkit, are useful for summarizing overall deviations across days and households.

@@

\caption{Scatterplots of half-hourly energy usage by time of day, with hourly averages overlaid, faceted by household and type of day. All households have different week days versus weekends daily routines. On week days, household 1 wakes up early before 6am, and household 2 around 6am, followed by household 3 and 4. The use of air conditioning is notable in households 1 and 2, as seen by horizontal clusters of points.}\label{fig:hod}
\end{figure}

@@

Figure \ref{fig:calendar-elec} displays the global scaling of each household's data in a calendar layout, unfolding their day-to-day life via electricity usage. Glancing over household 1, their overall energy use is relatively low. Their week day energy use is distinguishable from their weekends, indicating a working household. The air conditioner appears to be used in the summer months (January and February) for a couple of hours in the evening and weekends. In contrast, household 2 keeps a cooling system functioning for much longer hours, which becomes more evident from late Wednesday through Thursday to early Friday in mid-January. These observations help to explain the stripes and clusters of household 2 in Figure \ref{fig:hod}. It is difficult to give a succinct description of household 3 since everyday energy pattern is variable, but May and June see more structure than the previous months. Individual data can be idiosyncratic, hence aggregated plots like Figure \ref{fig:dow} and \ref{fig:hod} are essential for assembling pieces to form a picture. However, the calendar plots tell the stories that are untold by previous plots, for example, their vacation time. Household 1 is on vacation over three weeks of mid-June, and household 2 also took some days off in the second week of June. Further, household 3 takes one short trip in January and the another one starting in the fourth week of June. Household 4 is away over two or three weeks in early April and late June. They all tend to take breaks during June probably due to the fact that the University winter break starts in June.

>>>>

\bigskip
\begin{abstract}
Calendars are broadly used in society to display temporal information and events. This paper describes a new calendar display for plotting data, that includes a layout algorithm with many options, and faceting functionality. The functions use modular arithmetic on the date variable to restructure the data into a calendar format. The user can apply the grammar of graphics to create plots inside each calendar cell, and thus the displays synchronize neatly with \textbf{ggplot2} graphics. The motivating application is studying pedestrian behavior in Melbourne, Australia, based on counts which are captured at hourly intervals by sensors scattered around the city. Faceting by the usual features such as day and month, is insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to events such as work days, weekends, holidays, and special events. The functions for the calendar algorithm are available in the R package \textbf{sugrrants}.
\end{abstract}

@@

A new method for organizing and visualizing temporal data, collected at sub-daily intervals, into a calendar layout is developed. The format is created using modular arithmetic, giving a restructuring of the data that can then be integrated into a data pipeline. The core component of the pipeline is to visualize the resulting data using the grammar of graphics \citep{wilkinson2006grammar, wickham2009ggplot2}, as used in \textbf{ggplot2} \citep{R-ggplot2}, where plots are defined as a functional mapping from variables in the data to graphical elements. The data restructuring approach is consistent with the tidy data principles available in the \textbf{tidyverse} suite of tools \citep{R-tidyverse}. The methods are implemented in a new R package called \textbf{sugrrants} \citep{R-sugrrants}.

The purpose of the calendar-based visualization is to provide insights into people's daily schedules, relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia \citep{ped}. There are many sensors installed across the inner-city area, that count pedestrians every hour (Figure \ref{fig:ped-map}). Data from 43 sensors in 2016 is analyzed here. This data can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. Patterns relative to special events (such as public holidays and recurring cultural/sporting events) would be worth studying in comparison to regular days, but conventional displays of time series data may bury this detail.

@@

\caption{Google map of the Melbourne city area, grey dots indicate sensor locations. The three locations highlighted will be analyzed in the paper: the State Library is a public library; Flagstaff Station is a train station, closed on non-work days; Birrarung Marr is an outdoor park hosting many cultural and sports events.}\label{fig:ped-map}
\end{figure}

@@

In contrast, calendar-based graphics unpack the temporal variable, at different resolutions, to digest multiple seasonalities and special events. There are some existing works in this area. For example, \citet{VanWijkCluster1999} developed a calendar view of the heatmap to represent the number of employees in the work place over a year, where colors indicate different clusters derived from the days. It contrasts weekdays and weekends, highlights public holidays, and presents other known seasonal variation such as school vacations, all of which have influence over the turn-outs in the office. \citet{Jones-heatmap}, \citet{R-TimeProjection}, \citet{R-ggTimeSeries}, and \citet{R-ggcal} implemented some variants of calendar-based heatmaps in R packages: \textbf{TimeProjection}, \textbf{ggTimeSeries}, and \textbf{ggcal} respectively. However, these techniques are limited to color-encoding graphics and are unable to use time scales smaller than a day. Time of day, which serves as one of the most important aspects in explaining substantial variations arising from the pedestrian sensor data, will be neglected through daily aggregation. Color-encoding is also low on the hierarchy of optimal variable mapping \citep{cleveland1984graphical, lam2007overview}.

The proposed algorithm goes beyond the calendar-based heatmap. The approach is developed with three conditions in mind: (1) to display time-of-day variation in addition to longer temporal components such as day-of-week and day-of-year; (2) to incorporate lines and other types of glyphs into the graphical toolkit for the calendar layout; (3) to accentuate unusual patterns, such as those related to special events, for viewers. The proposed algorithm has been implemented in the \texttt{frame\_calendar()} and \texttt{facet\_calendar()} functions in the \textbf{sugrrants} package using R.

@@

The algorithm of transforming data for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data \citep{Wickham2012glyph}. To make a year long calendar requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month conforms to a layout of 5 rows and 7 columns, where the top left is Sunday of week 1, and the bottom right is Monday of week 5 by default. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday--Sunday, which is deterministic given the year of the calendar. Months are of different lengths, ranging from 28 to 31 days. Some months could extend over six weeks, but for these months the last few days are wrapped up to the top row of the block for compactness, and because it is convention. The notation for creating these cells is as follows:

\begin{itemize}
\tightlist
\item
  \(k = 1, \dots , 7\) is the day of the week, that is the first day of the month.
\item
  \(d = 28, 29, 30\) or \(31\) representing the number of days in any month.
\item
  \((i, j)\) is the grid position where \(1 \le i \le 5\), is the row (week within the month), \(1 \le j \le 7\), is the column (day of the week), with \((1, 1)\) being in the upper left corner.
\item
  \(g = k, \dots,(k + d - 1)\) indexes the day in the month, inside the 35 possible cells.
\end{itemize}

The grid position for any day in the month is given by
\begin{equation}
  \begin{aligned}
  i &= \lceil (g \text{ mod } 35) / 7\rceil, \\
  j &= g \text{ mod } 7 + 1.
  \end{aligned}
  \label{eq:grid}
\end{equation}

\noindent Figure \ref{fig:month-diagram} illustrates this \((i, j)\) layout for a month where \(k = 5\).

@@

To create the layout for a full year, \((m, n)\) denotes the position of the month arranged in the plot, where \(1 \le m \le M\) is the row and \(1 \le n \le N\) is the column; \(b\) denotes the small amount of white space between each month for visual separation. Figure \ref{fig:year-diagram} illustrates this layout where \(M = 3\) and \(N = 4\).

@@

\caption{The calendar plot of hourly foot traffic at Flagstaff Station, using line graphs. The disparities between weekday and weekend along with public holiday, are immediately apparent. The arrangement of the data into a \(3 \times 4\) monthly grid represents all the traffic in 2016. Note that the algorithm wraps the last few days in the sixth week to the top row of each month block for a compact layout, which occurs in May and October.}\label{fig:fs-2016}
\end{figure}

Figure \ref{fig:fs-2016} shows the line graphs framed in the monthly calendar over the year 2016. This is achieved by the \texttt{frame\_calendar()} function, which computes the coordinates on the calendar for the input data variables. These can then be plotted using the usual \textbf{ggplot2} R package \citep{R-ggplot2}. Thus, the grammar of graphics can be applied.

In order to make calendar-based graphics more accessible and informative, reference lines dividing each cell and block, as well as labels indicating weekday and month are also computed before plot construction.

@@

The month labels located on the top left using \((\min{(x)}, \max{(y)})\) for every \((m, n)\). The weekday texts are uniformly positioned on the bottom of the whole canvas, that is \(\min{(y)}\), with the central position of a cell \(x / 2\) for each \(j\).

@@

It is assumed that the \texttt{data} is in tidy format \citep{wickham2014tidy}, and \texttt{x}, \texttt{y} are the variables that will be mapped to the horizontal and vertical axes in each cell. For example in Figure \ref{fig:fs-2016}, the \texttt{x} is the time of the day, and \texttt{y} is the count. The \texttt{date} argument specifies the date variable in the data, facilitating the range of dates plotted in the calendar layout.

@@

The start of the week for a monthly calendar is adjustable. The default is Monday (\texttt{1}), which is chosen from the data perspective. The week, however, can begin with Sunday (\texttt{7}) as commonly used in the US and Canada, or other weekday, subject to different countries and cultures.

@@

When \texttt{polar\ =\ TRUE}, a polar transformation is carried out on the data. The computation is similar to the one described in \citet{Wickham2012glyph}. This produces star glyphs \citep{chambers2017graphical}, where time series lines are transformed in polar coordinates, embedded in the monthly calendar layout. It is most useful in exhibiting cyclical patterns in the data.

@@

\caption{Line graphs on the calendar format showing hourly foot traffic at Flagstaff Station, scaled individually by day. The shape on a single day becomes more distinctive, as compared to Figure \ref{fig:fs-2016}.}\label{fig:fs-free}
\end{figure}

The \texttt{free\_wday} scales each weekday together. It can be useful to compare trends across weekdays, allowing relative patterns for weekends versus weekdays to be examined. Similarly, the \texttt{free\_mday} uses free scaling for any day within a given month.

@@

\caption{Line graphs, embedded in the \(6 \times 2\) monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasized.}\label{fig:facet}
\end{figure}

@@

Almost any type of plot can be shown in a calendar pane. Most of the full range of plotting capabilities in \textbf{ggplot2} is available. An example is shown in Figure \ref{fig:scatterplot}: the panes contain lag scatterplots, constructed with local scaling for each day at Flagstaff Station, where the lagged hourly count is assigned to the \texttt{x} argument and the current hourly count to the \texttt{y} argument. It indicates strong autocorrelation on weekends, and weak autocorrelation on work days. The V-shape in the weekday graphs arises when the next hour sees a substantial increase or decrease in counts. This is due to the morning and afternoon commuter patterns in Figure \ref{fig:fs-2016}: the adjacent hours are positively correlated when approaching to the peak hour but negatively correlated when moving away from the peak.

@@

\caption{Side-by-side boxplots of hourly counts for all the 43 sensors in December 2016, with the loess smooth line superimposed on each day. It shows the hourly distribution in the city as a whole. The increased variability is notable on the last day of December as New Year's Eve approaches. The month and weekday are labeled in Chinese, which demonstrates the support for languages other than English.}\label{fig:chn}
\end{figure}

@@

The previous calendar plots are static, made with \textbf{ggplot2}. The interactivity of calendar-based displays can be easily enabled, as long as the interactive graphics system remains true to the spirit of the grammar of graphics, for example, \textbf{plotly} \citep{plotly} in R. As a standalone display, an interactive tooltip can be added to show labels when mousing over a point in the calendar plot, for example the hourly count with the time of day. It is difficult to sense the values from the static display, but the tooltip makes it possible. Options in the \texttt{frame\_calendar()} function can be ported to a form of selection button or text input in a graphical user interface like R shiny \citep{R-shiny}. The display will update on the fly accordingly, via clicking or text input, as desired.

@@

Figure \ref{fig:dow} shows the energy use across days of week using boxplots. Inspecting the medians across households tells us that household 3, a family size of one couple and two children, uses more energy over the weekdays than other households. The relatively larger boxes for household 2 indicate greater variability in daily energy consumption with noticeable variations on Thursdays, and much higher usage over the weekends. The other two households (1 and 4) tend to consume more energy with more variation on the weekends relative to the weekdays, reflecting work and leisure patterns.

@@

\caption{Boxplots of daily energy usage against day of week for four households. Suggested by the medians, household 3 uses more energy than the others on the weekdays. By contrast, household 2 sees considerably larger variability.}\label{fig:dow}
\end{figure}

Figure \ref{fig:hod} shows energy consumption against time of day, separately by weekday and weekend. Household 1 is an early riser, starting their day before 6 and going back home around 18 on weekdays. They switch air conditioning on when they get home from work and keep it operating until midnight, evident from the small horizontal cluster of points around 0.8 kWh. On the other hand, the stripes above 1 kWh for household 2 indicates that air conditioning may run continuously for some periods, consuming twice the energy as household 1. A third peak occurs around 15 for household 3 only, likely coinciding when the children arrive home from school. They also have a consistent energy pattern between weekdays and weekends. As for household 4, their home routine starts after 18 on weekdays. Figures \ref{fig:dow} and \ref{fig:hod}, part of a traditional graphical toolkit, are useful for summarizing overall deviations across days and households.

@@

\caption{Scatterplots of half-hourly energy usage by time of day, with hourly averages overlaid, faceted by household and type of day. All households have different weekdays versus weekends daily routines. On weekdays, household 1 wakes up early before 6, and household 2 around 6, followed by household 3 and 4. The use of air conditioning is notable in households 1 and 2, as seen by horizontal clusters of points.}\label{fig:hod}
\end{figure}

@@

Figure \ref{fig:calendar-elec} displays the global scaling of each household's data in a calendar layout, unfolding their day-to-day life via electricity usage. Glancing over household 1, their overall energy use is relatively low. Their weekday energy use is distinguishable from their weekends, indicating a working household. The air conditioner appears to be used in the summer months (January and February) for a couple of hours in the evening and weekends. In contrast, household 2 keeps a cooling system functioning for much longer hours, which becomes more evident from late Wednesday through Thursday to early Friday in mid-January. These observations help to explain the stripes and clusters of household 2 in Figure \ref{fig:hod}. It is difficult to give a succinct description of household 3 since everyday energy pattern is variable, but May and June see more structure than the previous months. Individual data can be idiosyncratic, hence aggregated plots like Figure \ref{fig:dow} and \ref{fig:hod} are essential for assembling pieces to form a picture. However, the calendar plots tell the stories that are untold by previous plots, for example, their vacation time. Household 1 is on vacation over three weeks of mid-June, and household 2 also took some days off in the second week of June. Further, household 3 takes one short trip in January and another one starting in the fourth week of June. Household 4 is away over two or three weeks in early April and late June. They all tend to take breaks during June probably due to the fact that the University winter break starts in June.
